---
title: "Catchments Visualization"
author: "Sarah Popov"
date: "`r Sys.Date()`"
params:
  catchments: final_cc
  watersheds: watersheds_raw
  suitable_habitat: suitable_habitat
  cones: cones
  stn: stn
  nests: nests
  apikey: jawg_token
  headings: h_0
  h: h
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE)

# Load up the plotting functions
source("R/01-generate-radar-catchments.R")

sites <- params$catchments$site
#sites <- sites[1:2] # for testing
```

```{r test, eval = FALSE}
sites


params$catchments

params$watersheds

params$suitable_habitat

params$stn
```

Left plot = map of catchment for a given radar monitoring site with flight headings data

In red = watershed boundaries

In orange = cone containing the mean flight heading direction

Black points = known nest sites


Right plots = histogram of incoming (top plot) and outgoing (bottom plot) flight headings. The x-axis runs on a scale from 0-360°. The peak incoming and outgoing flight headings should be inverses of each other (peak incoming flights should == peak outgoing flights - 180°). 

Red line = bootstrapped mean incoming flight heading. 

Grey dashed lines = bootstrapped 95% confidence interval around mean.

```{r giant plot loop, eval = TRUE, include = TRUE, fig.width = 8.4, fig.height = 5.4}
plot_all <- function(site, 
                     catchments = params$catchments, 
                     watersheds = params$watersheds, 
                     suitable_habitat = params$suitable_habitat, 
                     cones = params$cones, 
                     stn = params$stn, 
                     nests = params$nests, 
                     apikey = params$apikey, 
                     headings = params$headings,
                     h = params$h) {
    p1 <- plot_final_map(site, 
                         catchments, 
                         watersheds, 
                         suitable_habitat, 
                         cones, 
                         stn, 
                         nests, 
                         apikey)
  p2 <- plot_headings(site, 
                      headings, 
                      h)
  p_all <- ggpubr::ggarrange(p1, p2, nrow = 1)
  return(p_all)
}

lapply(sites, function(i) {
  tryCatch({
    plot_all(site = i)
  },
  error = function(i) {
    message("Error with plot", i)
  }) # end tryCatch
}) # end lapply


# for (i in sites) {
# 
#   skip_to_next <- FALSE
# 
#   all_plots <- tryCatch({
#     plot_all(site = i)
#     }, 
#     error = function(e) { 
#       skip_to_next <<- TRUE
#     })
# 
#   if(skip_to_next) { next }     
# }
```

