---
title: "Nest Density distribution"
output: pdf_document
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warnings = FALSE)

library(targets)
library(sf)

withr::with_dir(rprojroot::find_root('_targets.R'), tar_load(nests))
withr::with_dir(rprojroot::find_root('_targets.R'), tar_load(sea_dist))
withr::with_dir(rprojroot::find_root('_targets.R'), tar_load(suitable_habitat))
```

# 1. Fit nest distribution curve

With nest distance-from-coast as the data, fit a spatial distribution curve that describes the probability density of finding a nest (Y) versus distance from the coast (X).

```{r cullen and frey, fig.cap = "Skewness-kurtosis plot of nest distances from the shore in km."}
library(fitdistrplus)

# Find the best distribution
descdist(nests$nest_dist_km, discrete = FALSE, boot = 1000)
```

```{r fit gamma distr, fig.cap = "Nest gamma distribution fit diagnostics. The top-left plot shows the empirical distribution of nest data (grey bars) overlaid with the fitted density curve (red line). The x-axis is distance from the coast in km. The remaining plots show model fit diagnostics."}
# Fit gamma distribution
fit <- fitdist(nests$nest_dist_km, distr = "gamma", method = "mle")

plot(fit)
```


\newpage

# 2. Spatialize the gamma distribution

Rasterize this curve, such that every pixel of the BC coastline runs from 0 (low/zero probability) to ~1 (high probability) of finding a Marbled Murrelet nest, based on the pixel's distance from the coast. 

```{r load sea dist, fig.cap = "Map with pixels colored by the distance from the coast in km."}
library(terra)

ifel(sea_dist > 0, sea_dist, NA) |> 
  plot(main = "Distance from coast (km)") # example coast distance raster
```

```{r rasterize gamma, fig.cap = "Rasterized gamma curve. Pixels are colored on a scale from 0 to 1, with higher numbers indicating a higher likelihood of nest occurrence. The value of the pixel is determined by the nest gamma distribution (Figure 2), with pixels closer to shore having a higher likelihood of nest occurrence. (Note that pixels immediately next to shore have a lower probability, with the peak in occurrence at around ~3 km from shore.)"}
# Replace distance from coast with gamma probabilities
# Derive a table of every distance with the gamma density 
# function value at each distance
distances <- terra::values(sea_dist)
# Predict values at each distance following the fit gamma distr 
g <- dgamma(distances, 
            shape = fit$estimate[[1]], 
            scale =  1 / fit$estimate[[2]])
# Normalize to 0-1
p <- g / max(g, na.rm = TRUE)

# Now put the nest probability values into a raster
p <- terra::rast(vals = p, 
                 crs = terra::crs(sea_dist), 
                 terra::ext(sea_dist), 
                 res = terra::res(sea_dist))

# Cut out ocean bits
p <- ifel(p == 0, NA, p)

# Plot
plot(p, main = "Nest probability")
```

Clip that nest probability raster to only occur within the suitable habitat layer.

```{r, echo = FALSE, fig.cap = "Rasterized gamma curve (Figure 4), clipped to Marbled Murrelet suitable habitat boundaries. This raster represents likelihood of nest occurrence."}
# Rasterize suitable habitat
temp <- terra::rast(terra::vect(sf::st_geometry(suitable_habitat)),
                    res = terra::res(sea_dist))
habitat <- terra::rasterize(terra::vect(suitable_habitat), temp)

habitat <- terra::resample(habitat, p) # resample `habitat` to be same extent as `p`

# multiply dist from coast nest probability raster * habitat raster
nest_prob <- habitat * p

plot(nest_prob, main = "Nest probability")
```

